---
title: "Annotation 0 - LN Unknown Clusters Exploration"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: cosmo
    highlight: pygments
    number_sections: true
    global_numbering: true  
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
---

# Introduction

Here the Visium LN niche annotation is explored with the objective of investigating Unknown clusters and label them appropiately.

```{r setup, include=FALSE}
set.seed(123)
source(here::here("misc/paths.R"))

knitr::knit_hooks$set(
  memory = function(before, options, envir) {
    if (!before) {
      gc(verbose = FALSE)
    }
  }
)
knitr::opts_chunk$set(memory = TRUE, echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      dev="svg",  out.width = "100%", fig.align = 'center',
                      
                      fig.path = here::here(paste0("{st_plots}" %>%
                                                     glue::glue() %>%
                                                     here::here(),
                                                   tools::file_path_sans_ext(basename(rstudioapi::getSourceEditorContext()$path)),
                                                   "_files/")
                                            )
                      )
options(width = 1200)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render(
  "0-annotation-ln.Rmd",
  output_format = NULL,
  output_dir = "{st_html}" %>%
                        glue::glue() %>%
                        here::here()
)
```

# Libraries

```{r}
library(Seurat)
library(ggsci)
library(cowplot)
library(patchwork)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db) 
```

# Load data

```{r}
se_ln <- "{st_dt}/se_annot_lymph-node.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
se_ln <- UpdateSeuratObject(se_ln)
```

# Report recap

In the LN annotation exploration from `3-report-st`, we saw that we have `r length(unique(se_ln$annot))` different annotated clusters, as we can see in Figure \@ref(fig:spannln). We can also see the proportion of each cluster for each sample in Figure \@ref(fig:propln).

```{r}
annot <- sort(unique(se_ln$annot))
palette <- pal_igv()(length(annot))
names(palette) <- annot
```

```{r spannln, fig.cap="Spatial feature plots showing the annotation across LN samples.", message=FALSE, warning=FALSE, dev='svglite', fig.height=8, fig.width=12}
p <- lapply(unique(se_ln$study_id), function(study_id) {
    SpatialDimPlot(
        object = se_ln[, se_ln$study_id == study_id],
        group.by = "annot",
        images = glue("{study_id}.lymph.node"),
        crop = FALSE,
        pt.size.factor = 1000,
        cols = palette) +
    labs(title = study_id) +
    theme(plot.title = element_text(hjust = 0.5, size = 18)) + 
    NoLegend()
})

dummy_data <- data.frame(x = 1:length(annot), y = 1:length(annot),
                         annot = factor(annot, levels = annot))

legend_plot <- ggplot(dummy_data, aes(x = x, y = y, color = annot)) +
  geom_point() +
  scale_color_manual(values = palette, breaks = annot) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(title = "annot"))
legend <- get_legend(legend_plot)

wrap_plots(p, ncol = 4) / 
  wrap_elements(legend) +
  plot_layout(heights = c(16, 4)) 
```

```{r propln, fig.cap="Bar plots showing the proportion of each cluster across lung samples."}
se_ln@meta.data %>%
  ggplot(aes(x = study_id, fill = annot)) +
  geom_bar(position = "fill") +
  labs(x = "Study ID", y = "Proportion") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = palette) +
  theme_void() +
  theme(
    axis.text = element_text(color = "black", size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title = element_text(color = "black", size = 11),
    axis.title.y = element_text(angle = 90),
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10)) + 
  guides(fill = guide_legend(ncol = 2))
```

We also noticed that **we have 4 unknown clusters**, so here we will continue investigating these clusters.
Across the clusters, Unknown-1 was really similar to `Mithocondrial` cluster and will need to be investigated alone. Importantly, as we can see in Figure \@ref(fig:vlnmsl), all of the other `Unknown` clusters (2-4) seem to be from muscle/fibro-related niches as we can see by their expression of muscle-related genes and also based on their distribution on tissue.

```{r}
"{st_dt}/ln-gene-dict.R" %>%
    glue::glue() %>%
    here::here() %>%
    source(file = .)
```

```{r vlnmsl, fig.cap="Violin plots of the expression of muscle genes of interest for clusters Unknown-2, Unknown-3, Unknown-4 and Muscle-fibro.", fig.width=6, fig.height=10}
VlnPlot(subset(se_ln, annot %in% c("Unknown-2", "Unknown-3", "Unknown-4", "Muscle-Fibro")),
        features = c("MYH11", "DES", "MYLK", "ACTG2", "ACTA2", "MYL9","TAGLN", "MGP"),
        group.by = "annot",
        pt.size = 0.1,
        alpha = 0.3,
        ncol = 2)
```

# DE analysis

Starting from this point, we will perform a one-vs-all DE analysis for `Unknown2–4` to identify genes that differentiate each cluster from the others.

```{r}
s_se <- subset(se_ln, annot %in% c("Unknown-2", "Unknown-3", "Unknown-4"))
markers <- FindAllMarkers(
  object = s_se,
  assay = "Spatial",
  only.pos = TRUE,
  min.pct = 0.1
)
table(markers$cluster)

sign <-  markers %>%
  filter(p_val_adj < 0.01 & abs(avg_log2FC) > log2(1.5))
table(sign$cluster)
```

## Volcano plots

```{r volcano, fig.cap="Volcano plot of marker genes for Unknown-2-4 clusters."}
volcano <- function(t) {
  pThreshold <- max(t$p_val[t$p_val_adj < 0.01])
  
  keyvals.colour <- ifelse(
    t$avg_log2FC <= -log2(1.5) & t$p_val < pThreshold, 'blue',      
    ifelse(t$avg_log2FC >= log2(1.5) & t$p_val < pThreshold, 'red', 
    'gray80'))                                                         
  
  names(keyvals.colour)[keyvals.colour == 'blue'] <- 'Downregulated'
  names(keyvals.colour)[keyvals.colour == 'red'] <- 'Upregulated'
  names(keyvals.colour)[keyvals.colour == 'gray80'] <- 'Not significant'

  EnhancedVolcano(t, lab = t$gene, xlim = c(0, 10),
                  x = 'avg_log2FC', y = 'p_val', pCutoff = pThreshold, FCcutoff = log2(1.5),
                  pointSize = 1.5, labSize = 3.0,
                  title = t$cluster, subtitle = "", colCustom = keyvals.colour,
                  legendPosition = 'right', legendLabSize = 10, legendIconSize = 4.0,
                  drawConnectors = TRUE, widthConnectors = 0.5, 
                  max.overlaps = 40)
}

for (clust in unique(sign$cluster) ) {
  print(volcano(markers %>% filter(cluster == clust)))
}
```

## Heatmap (top markers)

```{r hm, fig.cap="Heatmap showing top 10 marker genes for Unknown-2-4 clusters.", dev='svglite'}
top10 <- sign %>%
  group_by(cluster) %>%
  top_n(10, avg_log2FC)

DoHeatmap(s_se, features = top10$gene, label = FALSE) + 
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0)
```

## GO-BP enrichment

```{r go, fig.height=8, fig.cap="Dot plot of enriched GO-BP for Unknown-2-4 clusters.", dev='svglite'}
go_results <- list()
for (clust in unique(sign$cluster) ) {
  
  genes <- sign %>%
    filter(cluster == clust) %>%
    pull(gene)
  entrez <- bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  
  go <- enrichGO(
    gene = entrez$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff = 0.05
  )
  go_results[[clust]] <- go
  
  print(dotplot(go, showCategory = 15) + ggtitle(clust))
}
```

# Interpretation

```{r uspdist, fig.height=4, fig.width=12, fig.cap="Spatial plots showing the distribution of Unknown clusters across LN samples.", dev='svglite'}
SpatialDimPlot(s_se,
               group.by = "annot",
               pt.size.factor = 1000,
               ncol = 4)
```

## Unknown-2

Based on the results we can see that this group of spots present stromal/fibro signature.

```{r u2ge, fig.height=10, fig.width=8, fig.cap="Violin plots showing expression of genes of interest for Unknown-2 across Unknown clusters."}
u2_ge <- c(# Fibro
  "COL1A2", "FN1", "PDGFRA", "PDGFRB",
  # Inflammatory
  "POSTN", "THBS2", "HBEGF", "PTGS2",
  # Chemokine/immune 
  "CCL19", "CCL2", "CXCL12", "IL7",
  # Myeloid 
  "CD68", "TYROBP",
  # Hypoxia
  "VEGFA", "HIF1A")

VlnPlot(s_se, 
        features = u2_ge,
        split.by = "annot",
        pt.size = 0.1,
        alpha = 0.3,
        ncol = 4) 
```

## Unknown-3 

Based on the found marker genes and GO enrichment results, we have evidence of `Unknown-3` possibly being Proliferating/Cycling Cells:

 - We have cell cycle genes (E2F2, NCAPG, BUB1B, MYB) as DE
 - We have several mitotic-related processes and lymphocyte differentiation enriched

```{r u3ge, fig.height=10, fig.width=8, fig.cap="Violin plots showing expression of genes of interest for Unknown-3 across Unknown clusters."}
u3_ge <- c(# Cell cycle/proliferation
  "MKI67", "TOP2A", "CDK1", "CCNB1",
  # DNA replication
  "MCM2", "MCM5", "PCNA", "RRM2",
  # Mitotic machinery
  "AURKB", "BUB1", "CENPE", "KIF11",
  # Lymphocyte
  "CD3D", # T
  "CD79A", # B
  # Stem/progenitor
  "SOX4", "HMGB2")

VlnPlot(s_se, 
        features = u3_ge,
        split.by = "annot",
        pt.size = 0.1,
        alpha = 0.3,
        ncol = 4) 
```

```{r barphase, fig.cap="Bar plot showing the distribution of cell cycle phases across Unknown clusters."}
ggplot(s_se@meta.data, aes(x = annot, fill = Phase)) +
  geom_bar(position = "fill") +
  theme_classic() +
  labs(y = "Proportion") +
  scale_fill_brewer(palette = "Set1")
```

Based on `Unknown-3` spots being highly proliferative (most cells in S/G2M phases) as seen in Figure \@ref(fig:barphase) we could label this cluster as `Proliferation`. Also, we have both a strong B and kind of T cell identity.

## Unknown-4

As we saw previously, `Unknown-4` is most likely a muscle-fibro cluster and in this analysis it is confirmed by a strong muscle signature (CASQ2, KCNAB1, GJA5, ACAN) and muscle-related GO terms.

```{r u4ge, fig.height=10, fig.width=8, fig.cap="Violin plots showing expression of genes of interest for Unknown-4 across Unknown clusters."}
u4_ge <- c(#Fibro
  "TAGLN", "ACTA2", "MYL9", "TPM2",
  "COL1A2", "FN1", "PDGFRA", "PDGFRB",
  # Muscle
  "LPP", "SYNM", "SORBS1", "NEXN",
  "POSTN", "THBS2",
  # Cell adhesion/ECM
  "ITGA5", "VCAN")

VlnPlot(s_se, 
        features = u4_ge,
        split.by = "annot",
        pt.size = 0.1, 
        ncol = 4,
        alpha = 0.3) 
```

# Unknown-1 cluster

Now we are going to repeat the enrichment but using the markers obtained from running `FindAllMarkers()` for all the clusters, so we can confirm our findings from clusters `Unknown2-4` and inspect `Unknown-1` too. 

## GO-BP enrichment

```{r goall, fig.height=10, fig.cap="Dot plot of enriched GO-BP for Unknown-1-4 clusters."}
markers_all <- "{st_dt}/mgs-lymph-node.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
markers_all <- subset(markers_all, cluster %in% c("3", "5", "8", "13"))
table(markers_all$cluster)

sign_all <-  markers_all %>%
  filter(p_val_adj < 0.01 & abs(avg_log2FC) > log2(1.5))
table(sign_all$cluster)

go_results_all <- list()
for (clust in unique(sign_all$cluster) ) {
  
  genes <- sign_all %>%
    filter(cluster == clust) %>%
    pull(gene)
  entrez <- bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  
  go <- enrichGO(
    gene = entrez$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff = 0.05
  )
  go_results_all[[clust]] <- go
  
  print(dotplot(go, showCategory = 15) + ggtitle(clust))
}
```

## Interpretation

As we mentioned before, the cluster `Unknown-1` (3) patterns of expression of the set of genes of interest are really similar to the `Mitochondial` cluster (1). This is confirmed by the results of the enrichment as the cluster shows nuclear/RNA processing functions and some mitochondrial metabolism. As we can see in Figure \@ref(fig:vlnmitou1) below, both clusters have really similar levels of mitochondrial expression.

```{r vlnmitou1, fig.cap="Violin plot showing percentage of MT- gene expression across all clusters."}
VlnPlot(se_ln,
        features = "percent.mito",
        group.by = "Spatial_snn_res.0.75",
        pt.size = 0.1,
        alpha = 0.3) + 
  NoLegend() + 
  scale_fill_manual(values = ifelse(levels(se_ln$Spatial_snn_res.0.75) %in% c("1", "3"), "#E64B35", "grey80"))
```

Regarding `Unknown-2`, we now see really different BP enriched, mainly related to detoxification and a strong stress response. We also have extracellular organization related terms suggesting some stromal identity as we saw in Figure \@ref(fig:u2ge).

```{r, fig.height=5, fig.width=8, fig.cap="Violin plots showing expression of genes of interest for Unknown-2 across Unknown clusters."}
u2_ge <- c(# Stress 
  "HSPA1A", "HSPB1", "GSTP1",
  # Metallothioneins
  "MT1X", "MT1A", "MT1E", "MT1M")

VlnPlot(s_se, 
        features = u2_ge,
        split.by = "annot",
        pt.size = 0.1,
        ncol = 4) 
```

Regarding `Unknown-3` and `-4`, we can be pretty sure about the previous interpretation, confirming that these consists of proliferating cells and muscle/fibro cells respectively.

# Annotation

```{r}
# re-label unknown clusters
```

# Save data

```{r, eval=FALSE}
annotation_dir <- glue("{st_odt}/1-annotation") %>% here::here()
dir.create(annotation_dir, recursive = TRUE, showWarnings = FALSE)

glue("{annotation_dir}/0-annotation-ln_mgs.rds") %>%
  saveRDS(object = markers, file = .)
glue("{annotation_dir}/0-annotation-ln_goBP_2-4.rds") %>%
  saveRDS(object = go_results, file = .)
glue("{annotation_dir}/0-annotation-ln_goBP_uall.rds") %>%
  saveRDS(object = go_results_all, file = .)
```

# Session Information

```{r}
sessionInfo()
```
