---
title: "Deconvolution 0 - Seurat data conversion to SummarizedExperiment classes"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: cosmo
    highlight: pygments
    number_sections: true
    global_numbering: true  
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
---
 
# Introduction

Here the snRNA-seq and visium data objects are converted to `SingleCellExperiment` and `SpatialExperiment`, that will be used as input for deconvolution with `SPOTlight`.

```{r setup, include=FALSE}
set.seed(123)
source(here::here("misc/paths.R"))

knitr::knit_hooks$set(
  memory = function(before, options, envir) {
    if (!before) {
      gc(verbose = FALSE)
    }
  }
)
knitr::opts_chunk$set(memory = TRUE, echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      dev="svg",  out.width = "100%", fig.align = 'center',
                      
                      fig.path = here::here(paste0("{st_plots}" %>%
                                                     glue::glue() %>%
                                                     here::here(),
                                                   tools::file_path_sans_ext(basename(rstudioapi::getSourceEditorContext()$path)),
                                                   "_files/")
                                            )
                      )
options(width = 1200)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render(
  "0-seurat_conversion.Rmd",
  output_format = NULL,
  output_dir = "{st_html}" %>%
                        glue::glue() %>%
                        here::here()
)
```

# Libraries

```{r}
library(Seurat)
library(SingleCellExperiment)
library(SpatialExperiment)
```

# snRNA-seq

## Load data

```{r}
sc <- "{sn_dt}/se_annot_consolidated.rds" %>%
    glue::glue() %>%
    here::here() %>%
    readRDS(.)
sc
```

## Conversion

```{r}
DefaultAssay(sc) <- "integrated" # if integrated will be used 
sce <- as.SingleCellExperiment(sc)
sce
# or 
# assay(sce, "integrated") <- assay(altExp(sce, "integrated"), "logcounts")
# to have all assays at the same level.
```

# Visium

## Load data

```{r}
st <- "{st_dt}/se_annot_join_lung.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
st <- UpdateSeuratObject(st)
```

## Conversion

An object consists of: **(i)** `assays` containing expression counts, **(ii)** `rowData` containing information on features, i.e. genes, **(iii)** `colData` containing information on spots or cells, including nonspatial and spatial metadata, **(iv)** `spatialCoords` containing spatial coordinates, and **(v)** `imgData` containing image data. It can also contain `reducedDims` where reduced dimensions as PCA, UMAP, ... are stored.

```{r}
as.SpatialExperiment <- function(se) {
  #do a loop for each sample? may not be necessary
  
  #spe <- SpatialExperiment(
  #  assays = list(counts = assay(sce)),
  #  rowData = rd, 
  #  colData = DataFrame(xyz), 
  #  spatialCoordsNames = c("pxl_col_in_fullres", "pxl_row_in_fullres"),
  #  imgData = img,
  #  sample_id = "foo")
  
  # matrix for the assays
  # DataFrame for coldata
}
```

spatialCoords can be supplied via colData by specifying the column names that correspond to spatial coordinates with spatialCoordsNames
(spatialCoordsNames = c("x", "y"))

Alternatively, spatialCoords may be supplied separately as a matrix 
(spatialCoords = xy)

EXAMPLE FUNC
```{r}
library(SpatialExperiment)
library(Seurat)
library(SeuratData)
# BiocManager::iBiocManager::install("SpatialExperiment")nstall("SpatialExperiment")
library(dplyr)

# SeuratData::InstallData(ds = "stxBrain")
SeuratData::LoadData(ds = "tonsilref")
se1 <- SeuratData::LoadData(ds = "stxBrain", type = "posterior1")
se2 <- SeuratData::LoadData(ds = "stxBrain", type = "anterior1")

## Function
seurat_to_spe <- function(seu, sample_id, img_id) {
    ## Convert to SCE
    sce <- Seurat::as.SingleCellExperiment(seu)
    
    ## Extract spatial coordinates
    spatialCoords <- as.matrix(
        seu@images[[img_id]]@coordinates[, c("imagecol", "imagerow")])
    
    ## Extract and process image data
    img <- SpatialExperiment::SpatialImage(
        x = as.raster(seu@images[[img_id]]@image))
    
    imgData <- DataFrame(
        sample_id = sample_id,
        image_id = img_id,
        data = I(list(img)),
        scaleFactor = seu@images[[img_id]]@scale.factors$lowres)
    
    # Convert to SpatialExperiment
    spe <- SpatialExperiment(
        assays = assays(sce),
        rowData = rowData(sce),
        colData = colData(sce),
        metadata = metadata(sce),
        reducedDims = reducedDims(sce),
        altExps = altExps(sce),
        sample_id = sample_id,
        spatialCoords = spatialCoords,
        imgData = imgData
    )
    # indicate all spots are on the tissue
    spe$in_tissue <- 1
    spe$sample_id <- sample_id
    # Return Spatial Experiment object
    spe
}

se_ls <- list("posterior1" = se1, "anterior1" = se2)
spe_ls <- lapply(names(se_ls), function(i) {
    seurat_to_spatialexperiment(seu = se_ls[[i]], sample_id = i, img_id = i)
})

spe <- Reduce(cbind, spe_ls)
spe
```


# Save data

```{r, eval=FALSE}
```

# Session Information

```{r}
sessionInfo()
```
