---
title: "Deconvolution 0 - Seurat data conversion to SummarizedExperiment classes"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: cosmo
    highlight: pygments
    number_sections: true
    global_numbering: true  
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
---
 
# Introduction

Here the snRNA-seq and visium data objects are converted to `SingleCellExperiment` and `SpatialExperiment`, that will be used as input for deconvolution with `SPOTlight`.

```{r setup, include=FALSE}
set.seed(123)
source(here::here("misc/paths.R"))

knitr::knit_hooks$set(
  memory = function(before, options, envir) {
    if (!before) {
      gc(verbose = FALSE)
    }
  }
)
knitr::opts_chunk$set(memory = TRUE, echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      dev="svg",  out.width = "100%", fig.align = 'center',
                      
                      fig.path = here::here(paste0("{st_plots}" %>%
                                                     glue::glue() %>%
                                                     here::here(),
                                                   tools::file_path_sans_ext(basename(rstudioapi::getSourceEditorContext()$path)),
                                                   "_files/")
                                            )
                      )
options(width = 1200)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render(
  "0-seurat_conversion.Rmd",
  output_format = NULL,
  output_dir = "{st_html}" %>%
                        glue::glue() %>%
                        here::here()
)
```

# Libraries

```{r}
library(Seurat)
library(SingleCellExperiment)
library(SpatialExperiment)
```

# snRNA-seq

## Load data

```{r}
sc <- "{sn_dt}/se_annot_consolidated.rds" %>%
    glue::glue() %>%
    here::here() %>%
    readRDS(.)
sc
```

## Conversion

```{r}
# if integrated will be used 

DefaultAssay(sc) <- "integrated"
sce <- as.SingleCellExperiment(sc)
sce

# or to have all assays at the same level.

# sce <- as.SingleCellExperiment(sc)
# assay(sce, "integrated") <- assay(altExp(sce, "integrated"), "logcounts")
# sce
```

# Visium

## Load data

```{r}
st <- "{st_dt}/se_annot_join_lung.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
st <- UpdateSeuratObject(st)
st
```

## Conversion

An object consists of: **(i)** `assays` containing expression counts, **(ii)** `rowData` containing information on features, i.e. genes, **(iii)** `colData` containing information on spots or cells, including nonspatial and spatial metadata, **(iv)** `spatialCoords` containing spatial coordinates, and **(v)** `imgData` containing image data. It can also contain `reducedDims` where reduced dimensions as PCA, UMAP, ... are stored.

```{r}
as.SpatialExperiment <- function(seu, sample_id, img_id) {
    sce <- as.SingleCellExperiment(seu) ## Convert to SCE
    
    ## Extract spatial coordinates
    spatialCoords <- as.matrix( 
        seu@images[[img_id]]@coordinates[, c("imagecol", "imagerow")])
    
    ## Extract and process image data
    img <- SpatialExperiment::SpatialImage( 
        x = as.raster(seu@images[[img_id]]@image))

    imgData <- DataFrame(
        sample_id = sample_id,
        image_id = img_id,
        data = I(list(img)),
        scaleFactor = seu@images[[img_id]]@scale.factors$lowres)
    
    ## Convert to SpatialExperiment
    spe <- SpatialExperiment(
        assays = assays(sce),
        rowData = rowData(sce),
        colData = colData(sce),
        metadata = metadata(sce),
        reducedDims = reducedDims(sce),
        altExps = altExps(sce),
        sample_id = sample_id,
        spatialCoords = spatialCoords,
        imgData = imgData
    )
    spe$in_tissue <- 1 # indicate all spots are on the tissue
    spe$sample_id <- sample_id # Return Spatial Experiment object
    spe
}

sample_ids <- names(st@images) # sample_ids <- gsub("\\.lung$", "", image_names)
se_ls <- lapply(sample_ids, function(id) {
    id_spots <- rownames(st@images[[id]]@coordinates)
    subset(st, cells = id_spots)
})
names(se_ls) <- sample_ids
```

Before we continue, lets make sure that we have splitted the Seurat object correctly.

```{r}
s_spots <- sapply(names(se_ls), function(sample) {
    ncol(se_ls[[sample]])
})
names(s_spots) <- gsub("\\.lung$", "", names(s_spots))
s_spots
table(st$study_id)
```

```{r}
spe_ls <- lapply(sample_ids, function(i) {
    as.SpatialExperiment(seu = se_ls[[i]], sample_id = i, img_id = i)
})
spe <- Reduce(cbind, spe_ls)
spe
```

# Save data

```{r, eval=FALSE}
```

# Session Information

```{r}
sessionInfo()
```
