---
title: "CellBender 1 - Review of CellBender output"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: cosmo
    highlight: pygments
    number_sections: true
    global_numbering: true  
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
---

# Introduction

Here the results obtained from running CellBender on the snRNA-seq raw dataset are analysed.

```{r setup, include=FALSE}
set.seed(123)
source(here::here("misc/paths.R"))

knitr::knit_hooks$set(
  memory = function(before, options, envir) {
    if (!before) {
      gc(verbose = FALSE)
    }
  }
)
knitr::opts_chunk$set(memory = TRUE, echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      dev="svg",  out.width = "100%", fig.align = 'center',
                      
                      fig.path = here::here(paste0("{st_plots}" %>%
                                                     glue::glue() %>%
                                                     here::here(),
                                                   tools::file_path_sans_ext(basename(rstudioapi::getSourceEditorContext()$path)),
                                                   "_files/")
                                            )
                      )
options(width = 1200)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render(
  "1-cellbender_review.Rmd",
  output_format = NULL,
  output_dir = "{st_html}" %>%
                        glue::glue() %>%
                        here::here()
)
```

# Libraries

```{r}
library(ggplot2)
library(patchwork) 
library(scCustomize)
library(Seurat)
library(SingleCellExperiment)
```

# Metrics summary

## Load data

```{r}
fn <- "{sn_odt}/0-cellbender" %>%
    glue::glue() %>%
    here::here() %>%
    list.files(path = ., full.names = TRUE) %>%
  list.files(path = ., full.names = TRUE, pattern = glue("metrics"))

metrics_ls <- lapply(fn, function(i) {
    tab <- read.table(file = i, sep = ",", col.names = c("metric", "value"))
    
    name <- basename(i)
    parts <- sub("^GSM[0-9]+_(.+)_raw_metrics\\.csv$", "\\1", name)
    split <- strsplit(parts, "-")[[1]]
    tab$id <- split[1]
    
    c_parts <- split[-1]
    c_parts <- c_parts[!grepl("^R[0-9]+$", c_parts)]
    tab$condition <- paste(c_parts, collapse = "-")
    
    return(tab)
})

metrics <- list_rbind(metrics_ls) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  arrange(condition, id)
metrics
```
## Metrics inspection

Summary statistics by condition

```{r}
metrics %>%
  group_by(condition) %>%
  summarise(
    n_samples = n(),
    mean_cells_found = mean(found_cells, na.rm = TRUE),
    mean_pct_removed = mean(fraction_counts_removed * 100, na.rm = TRUE),
    mean_counts_per_cell = mean(output_average_counts_per_cell, na.rm = TRUE),
    mean_convergence = mean(convergence_indicator, na.rm = TRUE),
    .groups = "drop"
  )
```

Total counts overview (raw, output, removed)
 
```{r}
palette <- setNames(c("#4AE38A", "#168D48","#1F78B4", "#0D542B")
                   [1:length(unique(metrics$condition[metrics$condition != ""]))],
                   unique(metrics$condition[metrics$condition != ""]))

metrics %>%
  select(id, condition, total_raw_counts, total_output_counts, total_counts_removed) %>%
  pivot_longer(cols = c(total_raw_counts, total_output_counts, total_counts_removed),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = factor(metric, levels = c("total_raw_counts", "total_output_counts", "total_counts_removed"))) %>%
  ggplot(aes(x = id, y = value, fill = condition)) +
  geom_col() +
  facet_wrap(~metric,  ncol = 1) + #scales = "free_y",
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = palette) +

  labs(title = "Total Counts Metrics", x = "Sample ID", y = "Counts")
```

proportion metrics

```{r}
metrics %>%
  select(id, condition, fraction_counts_removed, 
         fraction_of_analyzed_droplets_that_are_nonempty) %>%
  pivot_longer(cols = -c(id, condition), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = id, y = value, fill = condition)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_fill_manual(values = palette) +

  labs(title = "Fraction Metrics", x = "Sample ID", y = "Fraction")
```

Cell detection (expected, found, empties)

```{r}
metrics %>%
  select(id, condition, expected_cells, found_cells, found_empties) %>%
  pivot_longer(cols = c(expected_cells, found_cells, found_empties),
               names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = id, y = value, fill = metric)) +
  geom_col(position = "dodge") +
  facet_wrap(~condition, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Cell Detection: Expected vs Found vs Empties",
       x = "Sample ID", y = "Number of Droplets")
```

counts per cell 

```{r}
metrics %>%
  select(id, condition, output_average_counts_per_cell, average_counts_removed_per_cell,
         total_raw_counts_in_cells, total_counts_removed_from_cells) %>%
  pivot_longer(cols = -c(id, condition), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = id, y = value, fill = condition)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y", ncol = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = palette) +
  labs(title = "Per-Cell Count Metrics", x = "Sample ID", y = "Value")
```

Quality/convergence metrics

```{r}
 metrics %>%
  select(id, condition, convergence_indicator, overall_change_in_train_elbo, 
         ratio_of_found_cells_to_expected_cells) %>%
  pivot_longer(cols = -c(id, condition), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = id, y = value, fill = condition)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_fill_manual(values = palette) +
  labs(title = "Quality and Convergence Metrics", x = "Sample ID", y = "Value")
```

# Data comparison

## Load data

```{r}
fn <- "{sn_dt}" %>%
    glue::glue() %>%
    here::here() %>%
  list.files(path = ., full.names = TRUE, pattern = glue("h5"))


sample_names 

# Read all files
mat_ls <- lapply(seq_along(fn), function(i) {
  Read_CellBender_h5_Mat(file_name = h5_files[i])
})

names(mat_ls) <- sample_names

mat_merged <- Merge_Sparse_Data_All(matrix_list = mat_ls, add_cell_ids = names(mat_ls))
sce <- SingleCellExperiment(assays = list(counts = mat_merged))
sce


sce_ls <- lapply(fn, function(i) {
    tab <- Read10X_h5(i)
    sce <- SingleCellExperiment(assays = list(counts = as.matrix(tab)))
    
    return(sce)
})


h5_files <- list.files(path = here(glue("{sn_odt}/0-cellbender")),
                       pattern = "_raw_filtered\\.h5$",
                       recursive = TRUE, full.names = TRUE)
sample_names <- basename(dirname(h5_files))

# Read all files
mat_ls <- lapply(seq_along(h5_files), function(i) {
  Read_CellBender_h5_Mat(file_name = h5_files[i])
})

names(mat_ls) <- sample_names

mat_merged <- Merge_Sparse_Data_All(matrix_list = mat_ls, add_cell_ids = names(mat_ls))
sce <- SingleCellExperiment(assays = list(counts = mat_merged))
sce

# add metadata: donor, condition, nCount_RNA ?, nFeature_RNA ?. and:
#
sobj[['pct.mito']] <- PercentageFeatureSet(sobj, pattern='^MT-')
sobj[['pct.ribo']] <- PercentageFeatureSet(sobj, pattern='^RP[SL][0-9]*$')
sobj[['pct.MALAT1']] <- PercentageFeatureSet(sobj, pattern='^MALAT1$')
sobj[['pct.SCoV2']] <- PercentageFeatureSet(sobj, pattern='^SCoV2-')
```

class: SingleCellExperiment 
dim: 60675 86841 
metadata(0):
assays(1): counts
rownames(60675): DDX11L1 WASH7P ... SCoV2_ORF10 SCoV2_UTR3
rowData names(0):
colnames(86841): 17B_GAGGGTATCGATCCCT-1 17B_TCCTTTCGTTGGACCC-1 ... T188_TCGTGCTGTACTGGGA-1 T188_ATTTCACTCTCGTGGG-1
colData names(0):
reducedDimNames(0):
mainExpName: NULL

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```











