---
title: "Project report 1"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
    fig_caption: true
    keep_md: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev="svg",  out.width = "100%", fig.align = 'center', 
                      message = FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

This report contain the overview and exploration of the dataset and the analysis that have been previously performed for the project.

# Libraries

```{r}
library(Seurat)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(DataExplorer)
library(summarytools)
```

# Parameters

```{r}
set.seed(123)
```

# snRNA-seq

First, we are going to take a look at the snRNA-seq data after the pre-processing.

## Load data

```{r}
se <- "data/snRNAseq/lung_autopsy_all.rds" %>%
  here::here() %>%
  readRDS(.)
```

## Overview

In the **pre-processing**, the cells with more than 10% of of mitochondrial genes, more than 5,000 genes and doublets were removed. Then, the data was normalized using `NormalizeData()` function with a scale factor of 10,000 and the *LogNormalize* method. The 2,000 most variable features were identified using the `FindVariableFeatures()` function with the *vst* method, and cell cycle scores were calculated using the `CellCycleScoring()` function. Batch effect was then removed with `IntegrateData()` function using the donor as batch variable.
After scaling the data with the `ScaleData()` function, a PCA was performed using the `RunPCA()` function and the first 20 PCs were used to find neighbors with the `FindNeighbors()` function and to compute UMAP coordinates with the `RunUMAP()` function. Finally, clustering was performed using the `FindClusters()` function with the default parameters and the dataset by [Travaglini et al.](https://doi.org/10.1101/742320) was used to score cell types:

```{r}
DimPlot(se, group.by='predicted.id', label=TRUE, repel=TRUE) + NoLegend()
```

The data consists of `r length(unique(meta$orig.ident))` samples from different donors from lung autopsies, consisting of 6 COVID-19 cases and 2 controls. The data dimensions are `r ncol(se)` cells by `r nrow(se@assays$RNA@counts)` genes.

## Metadata descriptive analysis

```{r}
meta <- se@meta.data
plot_intro(meta)
head(meta)

# metadata of interest
categorical <- c("predicted.id", "Phase", "condition", "origin", "donor")
continuous <- c("prediction.score.max", "pct.mito", "pct.ribo", "pct.SCoV2",
  "nCount_RNA", "nFeature_RNA", "scrublet_score")
```

### Continuous variables

```{r}
descr(meta[,continuous])
plot_histogram(meta[,continuous], geom_histogram_args = list("fill" = "#a6cee3", "color" = "#1f78b4"), ggtheme = theme_minimal())
FeaturePlot(
    se,
    features = continuous)
```

### Categorical variables

```{r}
update_geom_defaults("bar", list(fill = "#a6cee3", color = "#1f78b4"))
plot_bar(meta[,categorical], ggtheme = theme_minimal())
DimPlot(
    se,
    group.by = categorical,
    label = FALSE) & NoLegend()
```


### SARS-CoV-2 positive cells
do descriptive analysis of scov2 by condition and cell type


```{r}
FeaturePlot(
    se,
    features = "pct.SCoV2", cells.highlight = WhichCells(se, expression = pct.SCoV2 > 0),
    cols = c("lightgrey", "red"),
    pt.size = 0.5) +
    labs(title = "SARS-CoV-2 positive cells")
```

```{r}
lapply(unique(se$condition), function(i) {
    VlnPlot(
        se[, se$condition == i],
        features = "pct.SCoV2",
        group.by = "predicted.id") +
        NoLegend() +
        labs(title = glue("{i} - pct.SCoV2"))
})
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

