---
title: "Project Report 2 - Visium Data & Analysis Exploration"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: cosmo
    highlight: pygments
    number_sections: true
    global_numbering: true  
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
---

# Introduction

This report contain the overview and exploration of the dataset and the analysis that have been previously performed for the project.

```{r setup, include=FALSE}
set.seed(123)
source(here::here("misc/paths.R"))

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,   tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      dev="svg",  out.width = "100%", fig.align = 'center',
                      
                      fig.path = here::here(paste0("{rep_plots}" %>%
                                                     glue::glue() %>%
                                                     here::here(),
                                                   tools::file_path_sans_ext(basename(rstudioapi::getSourceEditorContext()$path)),
                                                   "_files/")
                                            )
                      )
options(width = 1200)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render(
  "2-report-st.Rmd",
  output_format = NULL,
  output_dir = "{rep_html}" %>%
                        glue::glue() %>%
                        here::here()
)
```

# Libraries

```{r}
library(Seurat)
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(patchwork)
```

# Exploration

We are going to start exploring the Visium data and analysis that has been performed so far.
First, we are going to take a look at the data after the pre-processing and QC steps.

## Load data

We have the data divided by tissue and condition, we are going to load all the datasets in the same object to be able to explore it together.

```{r}
fn <- "{st_dt}/" %>%
    glue::glue() %>%
    here::here() %>%
    list.files(path = ., full.names = TRUE, pattern = glue("qc_se_"))
se_ls <- lapply(fn, function(i) {
    se <- readRDS(file = i)
})

se <- merge(
    se_ls[[1]], y = se_ls[2:length(se_ls)],
    project = "COVID")

meta <- se@meta.data
s_meta <- load_sample_metadata()
```

## Data overview

In the **QC** a basic analysis of the sequencing and mapping was performed from the QC metrics summary. Then, a QC visualization and filtering of Visium slides was done to prepare them for subsequent analysis. No-expressed genes by sample were removed, spots not overlapping tissue were also removed, and `chronic_case_4` from lymph node was removed due to low-quality. 

After this step, the data consists of `r length(unique(meta$cond_tissue))` samples from different donors from paired lung and lymph node autopsies, consisting of `r unique(meta$cond_tissue[grepl("lung", meta$cond_tissue)])` samples from lung and `r unique(meta$cond_tissue[grepl("lymph", meta$cond_tissue)])` from dLNs, from which we have a total of `r unique(meta$cond_tissue[grepl("control", meta$cond_tissue)])` controls. The data dimensions are `r ncol(se)` spots by `r nrow(se@assays$Spatial)` genes (after removing those genes that were’t expressed in any of the spots overlaying the tissue).
In Table \@ref(tab:exp), we can see a summary of the experimental design and the number of cells per donor:

```{r}
##### add tissue and remove n_cells:
meta
s_meta
```

```{r exp, tab.cap="Summary of the experimental design."}
tab <- s_meta %>%
  distinct(donor_id, disease_group, tissue) %>%
  pivot_wider(
    names_from = tissue,
    values_from = tissue,
    values_fill = ""
  ) %>%
  mutate(
    across(c(everything(), -donor_id, -disease_group), ~ifelse(. != "", "x", ""))
  ) %>%
  arrange(disease_group, donor_id) %>%
  rename(Donor = donor_id, Stage = disease_group, Lung = lung, `Lymph node` = `lymph-node`)

s_cols <- setNames(
  c("#4AE38A", "#168D48","#1F78B4", "#0D542B")[1:length(unique(tab$Stage[tab$Stage != ""]))],
  unique(tab$Stage[tab$Stage != ""])
)

kable(tab, format = "html", booktabs = TRUE, escape = FALSE, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE, 
                position = "center") %>%
  row_spec(6, color = "red") %>%
  column_spec(1, color = "#404040", bold = TRUE) %>%
  column_spec(2, color = s_cols[tab$Stage]) %>%
  column_spec(3, color = "black") 
```

## Metadata descriptive analysis
sample metadata!!!!!!!!!!

mirar sex, age
sars-cov-2_antemortem_swap
lung_fibrosis_score
tcr_id

## Basic data features exploration

We can visualize some basic featires of the data in figures x and y ...: 
Number of genes x spot
Number of reads x spot

```{r}
cond_tissue_ls <- s_meta %>%
    filter(!str_detect(string = s_meta$library_id, pattern = "\\*SP117_\\*")) %>%
    pull(study_id_tissue) %>%
    .[. != "chronic_case_4-lymph-node"]  

lapply(cond_tissue_ls, function(cond_tissue) {
  sample_id <- s_meta[s_meta$study_id_tissue == cond_tissue, ] %>% dplyr::pull(donor_id)
  
  Seurat::SpatialFeaturePlot(
    object = se[, se$cond_tissue == cond_tissue],
    features = "nFeature_Spatial",
    images = str_replace_all(cond_tissue, "-", ".")) +
      ggplot2::scale_fill_viridis_c(option = "magma") +
      ggplot2::labs(title = cond_tissue) +
      ggplot2::theme(plot.title = element_text(hjust = 0.5, size = 18))
  
}) %>% cowplot::plot_grid(plotlist = .)
```

######


tornar a fer SARS-CoV-2 % exploration

fer el mateix analysis (de l'estil) qe sn en general



######

*** notes on deconvolution
Notes tambe sobre tot el que necessito x la deconvolution, info computacional → que tenim, inputs que necessito, si ho puc correr a l’ordinador o ha de ser al cluster, …

deconvolution has to be done from 0 after `3-integration-clustering`. only for lung?

sn annotation is in `"{annot_sc}/{robj_dir}/se_annot_consolidated.rds"`
hi ha funcio de seurat a single cell exp pero no per st, sha de fer manualment


