---
title: "Project Report 2 - Visium Data & Analysis Exploration"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: cosmo
    highlight: pygments
    number_sections: true
    global_numbering: true  
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
---

# Introduction

This report contain the overview and exploration of the dataset and the analysis that have been previously performed for the project.

```{r setup, include=FALSE}
set.seed(123)
source(here::here("misc/paths.R"))

knitr::knit_hooks$set(
  memory = function(before, options, envir) {
    if (!before) {
      gc(verbose = FALSE)
    }
  }
)
knitr::opts_chunk$set(memory = TRUE, echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      dev="svg",  out.width = "100%", fig.align = 'center',
                      
                      fig.path = here::here(paste0("{rep_plots}" %>%
                                                     glue::glue() %>%
                                                     here::here(),
                                                   tools::file_path_sans_ext(basename(rstudioapi::getSourceEditorContext()$path)),
                                                   "_files/")
                                            )
                      )
options(width = 1200)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render(
  "2-report_st.Rmd",
  output_format = NULL,
  output_dir = "{rep_html}" %>%
                        glue::glue() %>%
                        here::here()
)
```

# Libraries

```{r}
library(Seurat)
library(kableExtra)
library(patchwork)
library(pheatmap)
library(colorBlindness)
library(ggsci)
library(cowplot)
library(DT)
```

# Integration and clustering

The integration and clustering of the data is done in 4 different stages:

 - First, harmony integration is done by condition and tissue to correct for batch effect. Here the data is normalized and scaled, PCA is computed on 3,000 HVG and SVG; then harmony is executed. Similarly, the data is then integrated at the tissue level.

 - In the next step the data is imputed using MAGIC denoising by tissue on selected marker genes for lung and LNs.

 - Finally, the clustering at different resolutions and annotation is done, also at the tissue level.

Now, we can explore the resulting data.

## Lung annotation

### Load data

```{r}
se_lung <- "{st_dt}/se_annot_join_lung.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
se_lung <- UpdateSeuratObject(se_lung)
```

### Exploration

In the case of lung samples, the annotation was previously done and used in a [paper](https://www.nature.com/articles/s41467-023-36333-2). However, the annotation was updated, we can see the changes in Figure \@ref(fig:confannot) below

```{r confannot, fig.cap="Confusion matrix of of the number of spots per cluster comparing the old and new annotation."}
se_lung@meta.data %>%
    count(newID, newID_old) %>%
    ggplot(aes(x = newID, y = newID_old, fill = n, label = n)) +
    geom_tile() +
    geom_text(color = "grey") +
    scale_fill_viridis_c(option = "viridis") +
      theme_classic() +
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
```



In the Figure \@ref(fig:dimintel) we can see how the data looks like after integration.

```{r dimintel, fig.cap="UMAPs showing the study ID and the annotation for lung samples.", fig.heigth=4, fig.width=12}
DimPlot(object = se_lung, 
        reduction = "umap", 
        group.by = c("study_id", "newID"), 
        pt.size = 0.1)
```

In Figure \@ref(fig:spotsxclustl) below, we can see the number of spots for each of the clusters for each sample.

```{r spotsxclustl, fig.cap="Confusion matrix of of the number of spots per cluster for each sample."}
se_lung@meta.data %>%
    count(newID, study_id) %>%
    ggplot(aes(x = newID, y = study_id, fill = n, label = n)) +
    geom_tile() +
    geom_text(color = "grey") +
    scale_fill_viridis_c(option = "viridis") +
      theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
```

Regarding the annotation, we have `r length(unique(se_lung$newID))` different annotated clusters, we can see them spatially distributed in Figure \@ref(fig:spannl). We can also see the proportion of each cluster for each sample in Figure \@ref(fig:propl)

```{r}
annot <- sort(unique(se_lung$newID))
palette <- paletteMartin[2:(length(annot)+1)]
names(palette) <- annot
```

```{r spannl, fig.cap="Spatial feature plots showing the annotation across lung samples.", message=FALSE, warning=FALSE, dev='svglite', fig.height=12, fig.width=12}
s_spots <- c("control_case_2-lung"," control_case_3-lung", "acute_case_1-lung",
             "chronic_case_2-lung", "chronic_case_4-lung",
             "prolonged_case_3-lung", "prolonged_case_4-lung") 

lapply(unique(se_lung$study_id), function(study_id) {
    pt_size <- ifelse(study_id %in% gsub("-lung", "", s_spots), 600, 1000)
    
    SpatialDimPlot(
        object = se_lung[, se_lung$study_id == study_id],
        group.by = "newID",
        images = glue("{study_id}.lung"),
        crop = FALSE,
        pt.size.factor = pt_size,
        cols = palette) +
      labs(title = study_id) +
      theme(plot.title = element_text(hjust = 0.5, size = 18))
}) %>% 
  wrap_plots(., ncol = 4) +
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```

```{r propl, fig.cap="Bar plot showing the proportion of each cluster across lung samples."}
se_lung@meta.data %>%
  ggplot(aes(x = study_id, fill = newID)) +
  geom_bar(position = "fill") + 
  labs(x = "Study ID", y = "Proportion") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = palette) +
  theme_void() +
  theme(
    axis.text = element_text(color = "black", size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title = element_text(color = "black", size = 11),
    axis.title.y = element_text(angle = 90),
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10))
```

### % SARS-CoV-2 exploration

Regarding SARS-CoV-2 gene expression as we can see in Table \@ref(tab:tabcovl) all clusters have spots with COVID reads.

```{r tabcovl, echo=FALSE, tab.cap = "Number of spots with SARS-CoV-2 gene expression by cluster."}
tab <- table(se_lung$percent.SCoV2 > 0, se_lung$newID)
kable(tab, format = "html", booktabs = TRUE, escape = FALSE) %>%
  scroll_box(width = "100%") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = TRUE, 
                position = "center")
```

```{r violin1l, fig.cap="Violin plots of the distribution of the percentage of SARS-CoV-2 gene expression by cluster for lung samples."}
VlnPlot(se_lung, features = "percent.SCoV2", group.by = "newID", pt.size = 0.1) + 
  scale_y_continuous(trans = "log1p") +
  xlab(NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend()
```

```{r violin2l, fig.cap="Violin plots of the distribution of the percentage of SARS-CoV-2 gene expression by cluster and stage for lung samples."}
VlnPlot(se_lung, features = "percent.SCoV2", group.by = "disease_group", split.by = "newID", pt.size = 0.1) + 
  scale_y_continuous(trans = "log1p") +
  xlab(NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend()
```

```{r, echo=FALSE}
rm(se_lung)
rm(annot)
rm(palette)
rm(tab)
```

## LNs annotation

### Load data

```{r}
se_ln <- "{st_dt}/se_annot_lymph-node.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
se_ln <- UpdateSeuratObject(se_ln)
```

### Exploration

In the case of LNs samples, in the Figure \@ref(fig:diminteln) we can see how the data looks like after integration.

```{r diminteln, fig.cap="UMAPs showing the study ID and the annotation for LN samples.", fig.heigth=4, fig.width=12}
DimPlot(object = se_ln, 
        reduction = "umap", 
        group.by = c("study_id", "annot"), 
        pt.size = 0.1)
```

In Figure \@ref(fig:spotsxclustln) below, we can see the number of spots for each of the clusters for each sample.

```{r spotsxclustln, fig.cap="Confusion matrix of the number of spots per cluster for each sample."}
se_ln@meta.data %>%
    count(annot, study_id) %>%
    ggplot(aes(x = annot, y = study_id, fill = n, label = n)) +
    geom_tile() +
    geom_text(color = "grey") +
    scale_fill_viridis_c(option = "viridis") +
      theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
```

Regarding the annotation, we have `r length(unique(se_ln$annot))` different annotated clusters, we can see them spatially distributed in Figure \@ref(fig:spannln). We can also see the proportion of each cluster for each sample in Figure \@ref(fig:propln)

**Note that we have 4 unknown clusters.**

```{r}
annot <- sort(unique(se_ln$annot))
palette <- pal_igv()(length(annot))
names(palette) <- annot
```

```{r spannln, fig.cap="Spatial feature plots showing the annotation across LN samples.", message=FALSE, warning=FALSE, dev='svglite', fig.height=8, fig.width=12}
p <- lapply(unique(se_ln$study_id), function(study_id) {
    SpatialDimPlot(
        object = se_ln[, se_ln$study_id == study_id],
        group.by = "annot",
        images = glue("{study_id}.lymph.node"),
        crop = FALSE,
        pt.size.factor = 1000,
        cols = palette) +
    labs(title = study_id) +
    theme(plot.title = element_text(hjust = 0.5, size = 18)) + 
    NoLegend()
})

dummy_data <- data.frame(x = 1:length(annot), y = 1:length(annot),
                         annot = factor(annot, levels = annot))

legend_plot <- ggplot(dummy_data, aes(x = x, y = y, color = annot)) +
  geom_point() +
  scale_color_manual(values = palette, breaks = annot) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(title = "annot"))
legend <- get_legend(legend_plot)

wrap_plots(p, ncol = 4) / 
  wrap_elements(legend) +
  plot_layout(heights = c(16, 4)) 
```

```{r propln, fig.cap="Bar plot showing the proportion of each cluster across LN samples."}
se_ln@meta.data %>%
  ggplot(aes(x = study_id, fill = annot)) +
  geom_bar(position = "fill") + 
  labs(x = "Study ID", y = "Proportion") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = palette) +
  theme_void() +
  theme(
    axis.text = element_text(color = "black", size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.title = element_text(color = "black", size = 11),
    axis.title.y = element_text(angle = 90),
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10)) +
  guides(fill = guide_legend(ncol = 2))
```

### % SARS-CoV-2 exploration

Regarding SARS-CoV-2 gene expression, we can see in Table \@ref(tab:tabcovln) from which cluster are the spots with COVID reads.

```{r tabcovln, echo=FALSE, tab.cap = "Number of spots with SARS-CoV-2 gene expression by cluster."}
tab <- table(se_ln$percent.SCoV2 > 0, se_ln$annot)
kable(tab, format = "html", booktabs = TRUE, escape = FALSE) %>%
  scroll_box(width = "100%") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = TRUE, 
                position = "center")
```

```{r violin1ln, fig.cap="Violin plots of the distribution of the percentage of SARS-CoV-2 gene expression by cluster for LN samples."}
VlnPlot(se_ln, features = "percent.SCoV2", group.by = "annot", pt.size = 0.1) + 
  scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 0.0001)) +
  xlab(NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend()
```

```{r violin2ln, fig.cap="Violin plots of the distribution of the percentage of SARS-CoV-2 gene expression by cluster and stage for LN samples."}
VlnPlot(se_ln, features = "percent.SCoV2", group.by = "disease_group", split.by = "annot", pt.size = 0.1) + 
  scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 0.0001)) +
  xlab(NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  NoLegend()
```

### Unknown clusters exploration

We are going to explore gene expression and spatial distribution of clusters labelled as `Unknown-x`.

```{r}
mgs <- "{st_dt}/mgs-lymph-node.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
head(mgs)

"{st_dt}/ln-gene-dict.R" %>%
    glue::glue() %>%
    here::here() %>%
    source(file = .)
```

```{r hm, fig.height=10, fig.cap="Heatmap showing average expression per cluster of the genes of interest."}
avg_expr <- AverageExpression(se_ln, features = gene_vec, group.by = "annot", assays = DefaultAssay(se_ln))
mat <- avg_expr[[DefaultAssay(se_ln)]]


pheatmap(mat, cluster_rows = TRUE, cluster_cols = TRUE,  scale = "row",
         border_color = "NA", 
         color = colorRampPalette(c("darkblue","blue", "white", "red", "darkred"))(256),
         fontsize_row = 10, fontsize_col = 12)
```

We can try to score the `Unknown` clusters, based on the marker genes from each one and the grouped genes of interest. We can see the scores for all the clusters in Figure \@ref(fig:hmscore)

```{r hmscore, fig.cap="Heatmap of the scores for the clusters based on groups of genes of interest."}
calculate_celltype_scores <- function(seurat_obj, gene_dict) {
  for (celltype in names(gene_dict)) {
    score_name <- paste0(gsub("[^[:alnum:]]", "_", celltype), "_score")
    
    genes_present <- gene_dict[[celltype]][gene_dict[[celltype]] %in% rownames(seurat_obj)]
    
    if (length(genes_present) > 0) {
      seurat_obj <- AddModuleScore(
        object = seurat_obj,
        features = list(genes_present),
        name = score_name,
        search = TRUE
      )
    }
  }
  return(seurat_obj)
}
se_ln <- calculate_celltype_scores(se_ln, gene_dict)

score_cols <- grep("_score", colnames(se_ln@meta.data), value = TRUE)
score_cols <- score_cols[score_cols != "lung_fibrosis_score"]

score_summary <- se_ln@meta.data %>%
  group_by(annot) %>%
  summarise(across(all_of(score_cols), mean, na.rm = TRUE)) %>%
  pivot_longer(cols = -annot, names_to = "celltype", values_to = "score") %>%
  mutate(celltype = gsub("_score.*", "", celltype))

ggplot(score_summary, aes(x = celltype, y = annot, fill = score)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Now we can see the highest average scores per cluster in Table \@ref(tab:assignment)

```{r assignment, echo=FALSE, tab.cap="Top 3 higher scores for the clusters based on groups of genes of interest."}
tab <- se_ln@meta.data %>%
  filter(grepl("Unknown", annot)) %>%
  group_by(annot) %>%
  summarise(across(all_of(score_cols), mean, na.rm = TRUE)) %>%
  pivot_longer(cols = -annot, names_to = "celltype", values_to = "score") %>%
  mutate(celltype = gsub("_score.*", "", celltype)) %>%
  group_by(annot) %>%
  slice_max(order_by = score, n = 3) %>%
  arrange(annot, desc(score))

kable(tab, format = "html", booktabs = TRUE, escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = TRUE, 
                position = "center")
```

#### Unknown-1 (cluster 3)

```{r sp1, fig.height=8, fig.width=12, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of cluster Unknown-1 across samples."}
u1 <- se_ln@meta.data %>%
    filter(annot == "Unknown-1") %>%
    pull(study_id) %>%
    unique()

lapply(u1, function(i) {
    SpatialDimPlot(
        se_ln[, se_ln$study_id == i & se_ln$annot == "Unknown-1"],
        images = glue("{i}.lymph.node"),
        crop = FALSE,
        pt.size.factor = 1000) +
    scale_fill_manual(values = "red") +
    labs(title = i) +
    NoLegend()
}) %>% wrap_plots(ncol = 4)
```

```{r, echo=FALSE, tab.cap="Marker genes of cluster Unknown-1."}
mgs_1 <- mgs %>%
  filter(cluster == "3") %>%
  arrange(desc(avg_log2FC))
mgs_mt <- mgs %>%
  filter(cluster == "1") %>%
  arrange(desc(avg_log2FC))

datatable(mgs_1)
```

From the Figure \@ref(fig:hm) we can see how the pattern of expression for the genes of interest of `Unknown-1` is really similar to `Mitochondrial`. If we take a look at the marker genes, these two clusters have `r sum(mgs_mt$gene %in% mgs_1$gene)` genes in common.

#### Unknown-2 (cluster 5)

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of cluster Unknown-2 across samples."}
u2 <- se_ln@meta.data %>%
    filter(annot == "Unknown-2") %>%
    pull(study_id) %>%
    unique()

lapply(u2, function(i) {
    SpatialDimPlot(
        se_ln[, se_ln$study_id == i & se_ln$annot == "Unknown-2"],
        images = glue("{i}.lymph.node"),
        crop = FALSE,
        pt.size.factor = 1000) +
    scale_fill_manual(values = "red") +
    labs(title = i) +
    NoLegend()
}) %>% wrap_plots(ncol = 4)
```

```{r, echo=FALSE, tab.cap="Marker genes of cluster Unknown-2."}
mgs_2 <- mgs %>%
  filter(cluster == "5") %>%
  arrange(desc(avg_log2FC))

datatable(mgs_2)
```

#### Unknown-3 (cluster 8)

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of cluster Unknown-3 across samples."}
u3 <- se_ln@meta.data %>%
    filter(annot == "Unknown-3") %>%
    pull(study_id) %>%
    unique()

lapply(u3, function(i) {
    SpatialDimPlot(
        se_ln[, se_ln$study_id == i & se_ln$annot == "Unknown-3"],
        images = glue("{i}.lymph.node"),
        crop = FALSE,
        pt.size.factor = 1000) +
    scale_fill_manual(values = "red") +
    labs(title = i) +
    NoLegend()
}) %>% wrap_plots(ncol = 4)
```

IDO depletes tryptophan and produces kynurenine, inducing T-cell apoptosis while simultaneously attracting Tregs and MDSCs (Abd El-Fattah 2022; Schupp et al. 2019)

IDO pathology staining - https://www.proteinatlas.org/ENSG00000131203-IDO1/tissue/lymph+node

```{r vlnu3, dev='svglite', fig.width=18, fig.height=15, fig.cap="Violin plots of of the expression of marker genes of interest in cluster Unknown-3, across all clusters."}
feat <- c(
    # Chemoattraction
    "CCL19", "CCL21", "CXCL13", # Interesting CXCL13 expression no?
    # Structural - Fibro
    "COL4A1", "PDGFRA", "PDGFRB",
    # ADAM is a transmembrane protein that contains a disintegrin and metalloprotease domain
    "ADAMTS2", "ADAMDEC1",
    # Serpin family
    "SERPINF2", "SERPINB9", "SERPINF1", "SERPINH1",
    # Interest
    "VEGFA", "IDO1",
    # MCM family
    "MCM2", "MCM4"
    )

lapply(feat, function(f){
  VlnPlot(
    se_ln,
    features = f,
    group.by = "annot",
    pt.size = 0.01) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
    NoLegend()
}) %>% wrap_plots(ncol = 4)
```

```{r, echo=FALSE, tab.cap="Marker genes of cluster Unknown-3."}
mgs_3 <- mgs %>%
  filter(cluster == "8") %>%
  arrange(desc(avg_log2FC))

datatable(mgs_3)
```

#### Unknown-4 (cluster 13)

```{r fig.height=8, fig.width=12, message=FALSE, warning=FALSE, fig.cap="Spatial distribution of cluster Unknown-4 across samples."}
u4 <- se_ln@meta.data %>%
    filter(annot == "Unknown-4") %>%
    pull(study_id) %>%
    unique()

lapply(u4, function(i) {
    SpatialDimPlot(
        se_ln[, se_ln$study_id == i & se_ln$annot == "Unknown-4"],
        images = glue("{i}.lymph.node"),
        crop = FALSE,
        pt.size.factor = 1000) +
    scale_fill_manual(values = "red") +
    labs(title = i) +
    NoLegend()
}) %>% wrap_plots(ncol = 4)
```

```{r, echo=FALSE, tab.cap="Marker genes of cluster Unknown-4."}
mgs_4 <- mgs %>%
  filter(cluster == "13") %>%
  arrange(desc(avg_log2FC))

datatable(mgs_4)
```

From the Figure \@ref(fig:hmscore) we can see that clusters `Unknown-2` and `Unknown-4` have relatively high scores for the muscle genes of interest. We can see the distribution of these genes across the clusters in Figure \@ref(fig:vlnmsl).

```{r vlnmsl, fig.cap="Violin plots of the expression of muscle genes of interest for clusters Unknown-2, Unknown-4 and Muscle-fibro.", fig.width=6, fig.height=10}
VlnPlot(subset(se_ln, annot %in% c("Unknown-2", "Unknown-4", "Muscle-Fibro")),
        features = c("MYH11", "DES", "MYLK", "ACTG2", "ACTA2", "MYL9","TAGLN", "MGP"),
        group.by = "annot",
        pt.size = 0.1,
        ncol = 2)
```

# Next steps: Deconvolution

The next step to follow is to do the deconvolution on the integrated ST data from lung (after `3-integration-clustering` steps), using the snRNA-seq dataset. 
Deconvolution will be done with `SPOTlight`, the minimal unit of data required to run it are:

 - ST (sparse) matrix with the expression, raw or normalized, where rows = genes and columns = capture locations.

 - Single cell (sparse) matrix with the expression, raw or normalized, where rows = genes and columns = cells.
 
 - Vector indicating the cell identity for each column in the single cell expression matrix.
 
In the [SPOTlight tutorial](https://www.bioconductor.org/packages/release/bioc/vignettes/SPOTlight/inst/doc/SPOTlight_kidney.html) objects of class SpatialExperiment (SE) and SingleCellExperiment (SCE) are used as data inputs.

We are using Seurat objects instead of classes derived from `SummarizedExperiment`, so to do the deconvolution the best would be to convert the data:

 - `as.SingleCellExperiment()` from Seurat can be used to convert the snRNA-seq data
 - ST data needs to be converted manually

# Session Information

```{r}
sessionInfo()
```
