---
title: "Quality Control 0 - Visium QC"
author: "Clàudia Vicente"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    theme: cosmo
    highlight: pygments
    number_sections: true
    global_numbering: true  
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: inline
---

# Introduction

Here a basic QC analysis of the mapping metrics from Spaceranger 2.0.1 and of of Visium slides is performed.

```{r setup, include=FALSE}
set.seed(123)
source(here::here("misc/paths.R"))

knitr::knit_hooks$set(
  memory = function(before, options, envir) {
    if (!before) {
      gc(verbose = FALSE)
    }
  }
)
knitr::opts_chunk$set(memory = TRUE, echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, tidy.opts = list(width.cutoff = 60),
                      dev="svg",  out.width = "100%", fig.align = 'center',
                      
                      fig.path = here::here(paste0("{st_plots}" %>%
                                                     glue::glue() %>%
                                                     here::here(),
                                                   tools::file_path_sans_ext(basename(rstudioapi::getSourceEditorContext()$path)),
                                                   "_files/")
                                            )
                      )
options(width = 1200)
```

```{r, eval = FALSE, include = FALSE}
rmarkdown::render(
  "0-QC-st.Rmd",
  output_format = NULL,
  output_dir = "{st_html}" %>%
                        glue::glue() %>%
                        here::here()
)
```

# Libraries

```{r}
library(kableExtra)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(Seurat)
library(ComplexHeatmap)
library(patchwork)
```

# Metrics QC

We will start by exploring mapping and sequencing metrics.

## Load data

```{r}
meta <- load_sample_metadata() 
ids <- meta %>% 
  pull(donor_id) %>%                  
  unique()  

files <- list.files(path = "{st_dt}/QC_metrics/" %>%
  glue::glue() %>%
  here::here(), 
  pattern = "-metrics_summary\\.csv$", full.names = TRUE)

metrics <- lapply(files, function(f) {
  
  file_base <- basename(f)
  base_clean <- str_remove(file_base, "-metrics_summary\\.csv$")
  
  parts <- str_split(base_clean, "-", n = 2)[[1]]
  sample_id <- parts[1]
  tissue <- parts[2]  
  
  tmp_df <- read_csv(f, show_col_types = FALSE)
  tmp_df <- tmp_df %>%
    dplyr::mutate(
      sample_id = sample_id,
      tissue = tissue
    ) %>%
    dplyr::select(-c(`Sample ID`)) %>%
    dplyr::select(sample_id, tissue, everything())
  
  return(tmp_df)
}) %>%
  bind_rows()

colnames(metrics) <- colnames(metrics) %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("\\(|\\)", "")

metrics <- metrics %>%
  mutate(sample_id_tissue = glue("{sample_id}-{tissue}"))
```

## Sequencing 

As we can see in Figure \@ref(fig:q30), Q30 bases in percentage are really near to 1 for all samples indicating reliability for base calling.

```{r q30, fig.height=3, fig.width=10, fig.cap="Bar plots showing percentage of Q30 bases per sample."}
l <- c("Q30_Bases_in_Barcode", "Q30_Bases_in_RNA_Read", "Q30_Bases_in_UMI")
palette <- colorRampPalette(colors = brewer.pal(8, "Paired"))(2)
names(palette) <- unique(metrics$tissue)
metrics_long <- metrics %>%
  pivot_longer(cols = all_of(l), names_to = "Metric", values_to = "Value")

ggplot(metrics_long, aes(x = sample_id_tissue, y = Value, fill = tissue)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  ylim(0, 1) +
  facet_wrap(~Metric, labeller = labeller(Metric = c("Q30_Bases_in_Barcode" = "in Barcode", 
                                                     "Q30_Bases_in_RNA_Read" = "in RNA Read", 
                                                     "Q30_Bases_in_UMI" = "in UMI"))) +
  scale_fill_manual(values = palette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample ID & Tissue", y = "Q30 Bases (%)")
```

## Mapping 

Next, we will check the quality of the mapping step. We aim to obtain libraries with a high percentage of confidently mapped reads, and specially a high percentage of exonic reads, which correspond with gene expression or RNAs. The reads mapping to intergenic regions suggest contamination of ambient DNA, whereas reads mapping to intronic regions may come from pre-mRNAs or mature spliced isoforms that retain certain introns.

```{r map, fig.cap="Bar plots showing percentage reads mapped confindently per sample."}
l <- c("Reads_Mapped_Confidently_to_Genome",
       "Reads_Mapped_Confidently_to_Intergenic_Regions",
       "Reads_Mapped_Confidently_to_Intronic_Regions", 
       "Reads_Mapped_Confidently_to_Exonic_Regions")
metrics_long <- metrics %>%
  pivot_longer(cols = all_of(l), names_to = "Metric", values_to = "Value")

ggplot(metrics_long, aes(x = sample_id_tissue, y = Value, fill = tissue)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  ylim(0, 1) +
  facet_wrap(~Metric, labeller = labeller(Metric = c("Reads_Mapped_Confidently_to_Genome" = "to genome", 
                                                     "Reads_Mapped_Confidently_to_Intergenic_Regions" = "to intergenic regions", 
                                                     "Reads_Mapped_Confidently_to_Intronic_Regions" = "to intronic regions",
                                                     "Reads_Mapped_Confidently_to_Exonic_Regions" = "to exonic regions"))) +
  scale_fill_manual(values = palette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample ID & Tissue", y = "Reads mapped confidently (%)")
```

## Sequencing saturation and depth

Now, we will plot the number of detected genes as a function of depth (sequenced reads). As sequencing depth increases, more genes are detected, but this function reaches a plateau, whereby more sequenced reads does not result in more detected genes; therefore, at this point we assure we sequenced until saturation. More specifically, the sequencing saturation the fraction of confidently mapped, valid cell-barcode, valid UMI reads that had a non-unique (cell-barcode, UMI, gene).

As we can see in Figure \@ref(fig:libsize), library sizes range from approximately `r round(min(metrics$Number_of_Reads)/1e6)` million to `r round(max(metrics$Number_of_Reads)/1e6)` million reads, with considerable variation between samples. Also, lymph-node samples have generally greater library sizes.

```{r libsize, fig.cap="Bar plots showing library size per sample."}
ggplot(metrics, aes(x = sample_id_tissue, y = Number_of_Reads / 1e6, fill = tissue)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(values = palette) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample ID & Tissue", y = "Library size (in millions)")
```

The Figure \@ref(fig:sat) below shows that, most samples achieved 75-100% saturation regardless of tissue type, indicating that the sequencing depth was generally sufficient. Lymph node samples required higher read depths (150-300k reads/spot) compared to lung samples (50-200k reads/spot) to reach saturation, suggesting different cellular complexity or RNA content between tissues.

```{r sat, fig.cap="Scatterplot showing mean reads per spot by sequencing saturation per sample."}
ggplot(metrics, aes(x = Mean_Reads_per_Spot / 1e3, y = Sequencing_Saturation, color = tissue)) +
  geom_point() +
  ylim(0, 1) +
  scale_color_manual(values = palette) +
  geom_text_repel(aes(label = sample_id_tissue), size = 4) +
  theme_minimal() +
  labs(x = "Mean Reads per Spot (in thousands)", 
       y = "Sequencing Saturation") +
  NoLegend()
```

As we can see in Figure \@ref(fig:depthcell), lymph node samples consistently detected more genes per spot than lung samples (1000-2000 genes). Sample `S317-lung` is an outlier with exceptionally high gene detection despite moderate sequencing depth.

```{r depthcell, fig.cap="Scatterplot showing mean reads by median detected genes per spot per sample."}
ggplot(metrics, aes(x = Mean_Reads_per_Spot / 1e3, y = Median_Genes_per_Spot, color = tissue)) +
  geom_point() +
  scale_color_manual(values = palette) +
  geom_text_repel(aes(label = sample_id_tissue), size = 4) +
  theme_minimal() +
  labs(x = "Mean Reads per Spot (in thousands)", 
       y = "Median Detected Genes per Spot") +
  NoLegend()
```

Despite varying sequencing depths (`r round(min(metrics$Number_of_Reads)/1e6)`-`r round(max(metrics$Number_of_Reads)/1e6)`M reads), as we see in Figure \@ref(fig:depth) most samples plateau at detecting 20,000-25,000 total genes, suggesting comprehensive transcriptome coverage. The major outlier is `S317-lymph-node`, which detected only `r metrics$Total_Genes_Detected[metrics$sample_id_tissue == "S317-lymph-node"]` genes despite very high read counts—this may indicate a technical problem requiring investigation in the remaining QC.

```{r depth, fig.cap="Scatterplot showing mean reads by median detected genes per spot per sample."}
ggplot(metrics, aes(x = Number_of_Reads / 1e6, y = Total_Genes_Detected, color = tissue)) +
  geom_point() +
  scale_color_manual(values = palette) +
  geom_text_repel(aes(label = sample_id_tissue), size = 4) +
  theme_minimal() +
  labs(x = "Library size (in millions)", y = "Total Genes Detected") +
  NoLegend()
```

Overall, the data quality appears generally good with adequate saturation, though there's clear biological distinction between tissue types. The S317 samples show unusual patterns that warrant further QC investigation.

# Visium QC

Now, we will follow with the QC of Visium slides. This QC has already been done and the filtering includes:

- `STutility::InputFromTable()` a default parameters filtering of spots and genes
- Filtering of genes that aren’t expressed in any of the spots
- Check if there are any spots not overlapping tissue

## Load data

We have the data divided by tissue and condition, we are going to load all the datasets in the same object to be able to explore it together.

```{r}
fn <- "{st_dt}/" %>%
    glue::glue() %>%
    here::here() %>%
    list.files(path = ., full.names = TRUE, pattern = glue("qc_se_"))
se_ls <- lapply(fn, function(i) {
    se <- readRDS(file = i)
    se <- UpdateSeuratObject(se)
    return(se)
})

se <- merge(
    se_ls[[1]], y = se_ls[2:length(se_ls)],
    project = "COVID")
```

## QC metrics summary per sample

First, looking at the main QC metrics by sample, we see in Figure \@ref(fig:gen) and Table \@ref(tab:summ) normal distribution of `n_counts` and `n_features` across all samples. It is worth noting that the `prolonged_case_4-lymph-node` and `control_case_1-lung` samples have a high mitochondrial content, spiking toward 20% or even higher. Again, there are some outlier spots reaching up to 40% for ribosomal content, particularly in the `control_case_2-lymph-node` sample.

```{r summ, echo=FALSE, tab.cap="Summary statistics per sample of the main QC metrics."}
tab <- se@meta.data %>%
  group_by(cond_tissue) %>%
  summarise(
    study_id = dplyr::first(study_id),
    tissue = dplyr::first(tissue),
    disease_group = dplyr::first(disease_group),
    n_spots = n(),
    mean_nCount = mean(nCount_Spatial),
    median_nCount = median(nCount_Spatial),
    sd_nCount = sd(nCount_Spatial),
    mean_nFeature = mean(nFeature_Spatial),
    median_nFeature = median(nFeature_Spatial),
    sd_nFeature = sd(nFeature_Spatial),
    mean_percent_mito = mean(percent.mito),
    median_percent_mito = median(percent.mito),
    mean_percent_ribo = mean(percent.ribo),
    median_percent_ribo = median(percent.ribo),
    mean_percent_SCoV2 = mean(percent.SCoV2),
    median_percent_SCoV2 = median(percent.SCoV2),
    .groups = 'drop')

t_cols <- palette
s_cols <- setNames(c("#4AE38A", "#168D48","#1F78B4", "#0D542B")
                   [1:length(unique(tab$disease_group[tab$disease_group != ""]))],
                   unique(tab$disease_group[tab$disease_group != ""]))

kable(tab[,-1], format = "html", booktabs = TRUE, escape = FALSE,
      col.names = c("ID","Tissue","Stage","Nº Spots", 
                    rep(c("Mean", "Median", "SD"), 2),
                    rep(c( "Mean", "Median"), 3))) %>%
  scroll_box(width = "100%") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = TRUE, 
                position = "center") %>%
  column_spec(1, color = "#404040", bold = TRUE) %>%
  column_spec(2, color = ifelse(tab$tissue == "", "white", t_cols[tab$tissue]), bold = TRUE) %>%
  column_spec(3, color = ifelse(tab$disease_group == "", "white", s_cols[tab$disease_group]), bold = TRUE) %>%
  add_header_above(c(" " = 4, "Nº Counts" = 3 , "Nº Features" = 3, "% Mitochondrial" = 2, "% Ribosomal" = 2, "% S-CoV-2" = 2))
```

```{r gen, fig.height=10, fig.width=12, fig.cap="Violin plots showing the distribution of nº counts, nº features, percentage of mitochondrial and ribosomal genes per sample.", dev='svglite'}
plt <- function(y_var, y_label, log = F) {
  p <- ggplot(se@meta.data, aes(x = cond_tissue, y = {{ y_var }}, fill = disease_group)) +
    geom_violin(trim = FALSE) +
    geom_jitter(size = 0.1, alpha = 0.3) +
    theme_classic() +
    theme(axis.text.x = element_blank()) +
    labs(x = "Sample", y = y_label) +
    scale_fill_manual(values = c("#4AE38A", "#168D48","#1F78B4", "#0D542B")) +
    NoLegend()
  
  if (log == T) {
    p <- p + scale_y_log10()
  } else {
    p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
  }
  
  return(p)
}

p1 <- plt(nCount_Spatial, "UMI counts (Log 10)", T)
p2 <- plt(nFeature_Spatial, "Detected genes (Log 10)", T)
p3 <- plt(percent.mito, "Mitochondrial genes (%)")
p4 <- plt(percent.ribo, "Ribosomal genes (%)")

(p1 | p2) / (p3 | p4)
```

Comparing UMI Counts versus Detected Genes in Figure \@ref(fig:cvsf), we can intuit a strong linear correlation between both variables for all samples. This is the expected behavior for high-quality data. We also can notice that the spots with higher mitochondrial content, in `prolonged_case_4-lymph-node`, are also of low-quality, having low UMI counts and detected genes.

```{r cvsf, fig.height=10, fig.width=14, fig.cap="Scatter plots showing nº counts versus nº features per sample"}
ggplot(se@meta.data, aes(x = nCount_Spatial, y = nFeature_Spatial, color = percent.mito)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_wrap(~cond_tissue, scales = "free", ncol = 5) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_viridis_c() +
  theme_bw() +
  labs(x = "UMI counts (log10)", y = "Detected genes (log10)")
```

## Outliers analysis

Now, we will identify potential outlier samples. We will calculate z-scores for each metric, defined by: 

$$Z=\frac{X−μ}{σ}$$
where, 

 - **X** is the mean QC metric for a specific sample (e.g., `mean_nCount`).
 - **μ** is the mean of that QC metric across all samples in your tab data frame (e.g., `mean(mean_nCoun)`).
 - **σ** is the standard deviation of that QC metric across all samples.

This normalization will allow us to compare the performance of each sample on a common scale. A Z-score of ±2 means a sample is two standard deviations above/below the average of all samples, we will use this SD theshold to consider a sample to be an outlier.

```{r}
outs_all <- tab %>%
  mutate(across(
    c(mean_nCount, mean_nFeature, mean_percent_mito, mean_percent_ribo, mean_percent_SCoV2),
    ~ (. - mean(.)) / sd(.),
    .names = "z_{.col}")) %>%
  rename_with(~ str_remove(., "mean_"), starts_with("z_mean_")) %>%
  mutate(low_umi = z_nCount < -2, high_umi = z_nCount > 2,
         low_genes = z_nFeature < -2, high_genes = z_nFeature > 2,
         high_mito = z_percent_mito > 2, high_ribo = z_percent_ribo > 2, high_SCoV2 = z_percent_SCoV2 > 2,
         is_outlier = low_umi | high_umi | low_genes | high_genes | high_mito | high_ribo | high_SCoV2)

outs_all %>%
  dplyr::select(cond_tissue, where(is.logical) & where(~any(., na.rm = TRUE))) %>%
  filter(is_outlier)
```

If we take a look at the outliers z-scores in Figure \@ref(fig:hm), we see that `acute_case_3-lung` is labelled as outlier due to high SCoV2 percentage, this could be expected as samples in acute stage should have the larger count of COVID-19 infection so we don't need to worry about it. Regarding `control_case_2-lymph-node` sample, it has the larger % Ribo value. We should worry the most about `prolonged_case_4-lymph-node` high mitochondrial content as it is `r round(outs_all$z_percent_mito[outs_all$cond_tissue == "prolonged_case_4-lymph-node"])` SD above the average of all samples. 

```{r hm, fig.cap="Heatmap showing z-scores of QC metrics of interest per sample."}
z_mat <- outs_all %>%
  dplyr::select(cond_tissue, z_nCount, z_nFeature, z_percent_mito, z_percent_ribo, z_percent_SCoV2) %>%
  column_to_rownames("cond_tissue") %>%
  as.matrix()

colnames(z_mat) <- c("UMI Count", "Gene Count", "% Mito", "% Ribo", "% SCoV2")
ann <- outs_all %>%
  dplyr::select(cond_tissue, is_outlier) %>%
  mutate(
    is_outlier = factor(is_outlier, levels = c(FALSE, TRUE), labels = c("No", "Yes")) 
  ) %>%
  column_to_rownames("cond_tissue")
ann <- HeatmapAnnotation(Outlier = ann$is_outlier, col = list(Outlier = c("No" = "#b5cf6b", "Yes" = "#ce6dbd")))

Heatmap(t(z_mat),
        name = "Z-score",
        top_annotation = ann,
        cluster_rows = FALSE,
        cluster_columns = TRUE)
```

```{r bp, fig.height=4, fig.width=8, fig.cap="Boxplots showing the distribution of percentage of mitochondrial, ribosomal and SARS-CoV-2 genes for all spots. Pink diamonds indicate mean values for outlier samples, with pink points showing spots from outlier samples overlaid on the overall distribution (green boxplots)."}
plt <- function(y_var, y_label, highlight_sample) {
  out_ids <- outs_all %>% filter(is_outlier) %>% pull(cond_tissue)
  dat <- se@meta.data %>%
    mutate(out_sample = cond_tissue %in% out_ids)
  mean_val <- dat %>% 
    filter(cond_tissue == highlight_sample) %>% 
    pull(.data[[y_var]]) %>% 
    mean()
  
  ggplot(dat, aes(x = "", y = .data[[y_var]])) +
    geom_boxplot(outlier.shape = NA, fill = "#b5cf6b", color = "#819e33", alpha = 0.5) +
    theme_classic() +
    theme(axis.text.x = element_blank()) +
    geom_jitter(data = dat %>% filter(out_sample & cond_tissue == highlight_sample),
                color = "#de9bd1", alpha = 0.3, width = 0.3, size = 0.5) +
    stat_summary(data = dat %>% filter(cond_tissue == highlight_sample),
                 fun = mean, geom = "point", shape = 23, size = 3, fill = "#ce6dbd", color = "#ce6dbd") +
    annotate("text", x = 1, y = mean_val + 2, 
             label = paste0(highlight_sample, " = ", round(mean_val, digits = 2)), 
             hjust = 0.5, vjust = -0.5, size = 3, color = "#581d4d") +
    labs(y = y_label, x = "")
}

p1 <- plt("percent.mito", "% Mitochondrial", "prolonged_case_4-lymph-node")
p2 <- plt("percent.ribo", "% Ribosomal", "control_case_2-lymph-node")
p3 <- plt("percent.SCoV2", "% SARS-CoV-2", "acute_case_3-lung")
(p1 | p2 | p3)
```

The Figure \@ref(fig:featribo) shows that the high content ribosomal areas from `control_case_2-lymph-node` are not randomly scattered but clustered in distinct, localized regions. The data suggests that the tissue section contains regions with high metabolic and proliferative activity, which is consistent with the biology of structures like follicles. High ribosomal percentages in these active regions are a biological signal, not a technical QC concern.

```{r featribo, fig.cap="Spatial feature plots showing the percentage of ribosomal expression per spot across Control case 2 from lymph node."}
SpatialFeaturePlot(
    object = se[, se$cond_tissue == "control_case_2-lymph-node"],
    features = "percent.ribo",
    pt.size.factor = 1000
  ) +
    scale_fill_viridis_c(option = "magma") +
    theme(plot.title = element_text(hjust = 0.5, size = 12))
```

Regarding mitochondrial expression, as we see in Figures \@ref(fig:bp) and \@ref(fig:featmito) the vast majority of the tissue area from `prolonged_case_4-lymph-node` shows a lot of spots with mitochondrial expression over ~15%. Moreover, if we plot the number of features and counts of the `prolonged_case_4` sample we see that mitochondrial expression is really indicating low quality of spots.

```{r featmito, fig.width=12, fig.cap="Spatial feature plots showing the number of features and counts per spot across Prolonged case 2 from lymph node."}
p1 <- lapply(c("prolonged_case_4-lymph-node"), function(cond_tissue) {
  sub <- se[, se$cond_tissue == cond_tissue & se$percent.mito < 30]

  SpatialFeaturePlot(object = sub[, sub$cond_tissue == cond_tissue],
                     features = "percent.mito",
                     pt.size.factor = 1000) +
    scale_fill_viridis_c(option = "magma")
    
})[[1]] 
p2 <- SpatialFeaturePlot(object = se[, se$cond_tissue == "prolonged_case_4-lymph-node"],
                     features = "nFeature_Spatial",
                     pt.size.factor = 1000) +
    scale_fill_viridis_c(option = "magma")
p3 <- SpatialFeaturePlot(object = se[, se$cond_tissue == "prolonged_case_4-lymph-node"],
                     features = "nCount_Spatial",
                     pt.size.factor = 1000) +
    scale_fill_viridis_c(option = "magma")
(p1 | p2 | p3)
```

If we take a look at the LN integrated data we see that `prolonged_case_4` is problematic as almost all its spots are clustered together.

```{r, fig.heigth=4, fig.width=18, fig.cap="UMAPs showing the mitochondrial percentage, the study ID and the annotation for LN samples."}
rm(list = setdiff(ls(), "st_dt"))
se_ln <- "{st_dt}/se_annot_lymph-node.rds" %>%
    glue::glue() %>%
    here::here() %>%
  readRDS(file = .)
se_ln <- UpdateSeuratObject(se_ln)
p1 <- FeaturePlot(se_ln, features = "percent.mito")
p2 <- DimPlot(object = se_ln, reduction = "umap", group.by = c("study_id",
    "annot"), pt.size = 0.1)
(p1 | p2) + plot_layout(widths = c(1, 2))
```

# Session Information

```{r}
sessionInfo()
```
